Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Final State with Multivariate Techniques

### 2015-12-08 paper/sec-mva.tex
More work on paper

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
  \begin{center}
      \vspace{-1cm}
  \includegraphics[width=0.90\textwidth]{plots/bst_nnarch_noPU.pdf}
  \vspace{-2cm}
  \caption{\small Architecture of the Artificial
    Neural Network (ANN)
    used for the analysis of the
    boosted
    category, with $N_{\rm var}=21$ input variables and thus
    the same number of neurons
  in the first layer.
  %
  The color code in the neuron connections (the weights) is a heat map,
  with red indicating large values and black small values.
}
\label{fig:nnarch}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

The training of the ANN for the signal/background classification task
proceeds as follows.
%
Given a set of $N_{\mathrm{var}}$  kinematic variables $\{k\}_i$ associated with the event $i$, and a set of neural network parameters $\{\omega\}$, we interpret the neural network output $y_i$ (that is, the activation state of the
neuron in the last layer)
as the probability that the event $i$ originates from the signal process,
\be
y_i = P(y^\prime_i=1|\{k\}_i, \{\omega\} )\, ,
\ee
where $y_i^\prime$ represents the true classification of the event $i$, i.e $y^\prime_{\text{signal}} = 1$ and $y^\prime_{\text{background}} = 0$. With this interpretation, our general classification probability including background events is given by
\be
P(y_i^\prime|\{k\}_i, \{\omega\}) = y_i^{y^\prime_i}(1-y_i)^{1-y^\prime_i} \, ,
\ee
which implies that the  negative log-likelihood cost function
that needs to be minimized during the ANN training is 
the cross-entropy error, defined as
 \bea
 E(\{\omega\}) &=& -\log\left(\prod_i^{N_{\text{evt}}} P(y_i^\prime|\{k\}_i, \{\omega\})\right)\nonumber\\
 &=&
 \sum_i^{N_{\text{evt}}} y^\prime_i\log{y_i} + (1-y^\prime_i)\log{(1-y_i)} \, ,
 \label{cross-entropy}
 \eea
 where $N_{\text{evt}}$ is the number of
 Monte Carlo events that we have generated.
 %
 The ANN is trained both on the signal and background MC events,
 in each case using the information $y_i'$ of the true nature
 of the event.
 %
 To be stable against statistical fluctuations, it is important to use
 a large enough MC sample both of signal and of background events
 for the MVA training.

 Therefore, the training of the neural networks consist on the
 minimization of the cross-entropy error,
 Eq.~(\ref{cross-entropy}), which in this work is achieved using a
 Genetic Algorithm.
 %
 The GA training is performed for a very large 
 number of generations, in this case $N_{\rm gen}=50k$, to avoid
 under-training.
 %
 We have verified that if a larger number of generations
 are used the results are unchanged.
 %

@@

 To identify the variables with higher discriminatory power,
 we need to train
 the MVA using a large number of kinematic distributions for
 signal and background events.
%
In our $hh\to 4b$ case,
the input variables for the MVA differ between our three categories,
and in particular in the case of large-$R$ jets (for the boosted
and intermediate categories) we want to exploit
the richness of substructure information.

For the three categories, boosted, intermediate and resolved,
the following common variables are used as input to the MVA:
\begin{itemize}
\item The transverse momenta of the leading and subleading Higgs, $p_{T,h_1}$ and $p_{T,h_2}$.
\item The transverse momentum of the reconstructed Higgs pair, $p_{T,hh}$.
\item The invariant masses of the leading and sub-leading Higgs candidates, $m_{h,1}$ and $m_{h,2}$.
\item The invariant mass of the reconstructed Higgs pair, $m_{hh}$.
\item The separation in the $\phi$--$\eta$ plane
  between the two Higgs candidates, $\Delta R_{hh}$.
  \item The separation in $\eta$  between the two Higgs candidates, $\Delta \eta_{hh}$.
\item The separation in $\phi$  between the two Higgs candidates, $\Delta \phi_{hh}$.
\end{itemize}
In addition, in the boosted category we use
  the transverse momenta of the leading, $p_{T,h_{1,1}}$ and $p_{T,h_{1,2}}$ and
  sub-leading, $p_{T,h_{2,1}}$ and $p_{T,h_{2,1}}$, Higgs candidate subjets.
  %
  In the resolved category instead,
  the corresponding variables are
  the transverse momenta $p_{T,i}$ of the four leading 
  $b$-tagged small-$R$ jets in the event.
  %
  In the intermediate category, we use the
  transverse momenta of the subjets
  from the large-$R$ jet $p_{T,h_{1,1}}$ and $p_{T,h_{1,2}}$ and the
 transverse momenta $p_{T,i}$ of the two leading 
  $b$-tagged small-$R$ jets.
  %
Therefore, we have a total of 13 variables which are common to the three categories.
%
Therefore, there are in total $N_{\mathrm{var}}=13$ variables for the
resolved analysis,
17 for the intermediate,
and 21 for the boosted category.
%
Given that the MVA is able to identify the most discriminatory variables
and to ignore those that have smaller effect, it is advantageous to
include as many substructure variables as possible.
%
This is one of the main advantages of ANNs in this context: they are
inherently redundant, so
adding additional information, even if carries very little weight,
should not degrade
the classification power of the MVA (other than making the GA
minimization less efficient).

@@

In Fig.~\ref{fig:nnresponse} we show the distribution of
the ANN output at the end of the training, for the exclusive
boosted, intermediate and resolved categories.
%
The distributions are normalized so that their integral
  adds up to one.
%
The  separation between signal and background is achieving by introducing
a cut, $y_{\rm cut}$, in the ANN output, so that MC events with $y_i\ge
y_{\rm cut}$ are classified as signal events, and those with
 $y_i <
y_{\rm cut}$ as background events.
%
Therefore,
the more separated that the two distributions are, the more efficient
the MVA discrimination will be.

@@

As can be seen from Fig.~\ref{fig:nev2}, in the boosted category
event with a cut in the ANN output as hard as $y_{\rm cut}\simeq 0.9$
we are still left
with around 100 events at the HL-LHC, 
which the background being reduced substantially to around
1000 events.
%
A similar cut would work fine for the intermediate category, though here
the backgrounds are much larger and thus the expected signal significance smaller.
%
For the resolved case, the signal significance would be only slightly
improved for any value of the cut.

A useful property of MVAs such as the one used in this work
is that they provide direct  physical insight about which of the
input variables contribute the most to the separation of
signal over background.
%
In the case of ANNs, this can be quantified by computing the sum
of the absolute value of al the weights connected to a given
input neuron $i$, that is
\be
\label{eq:totweight}
\omega^{\rm tot}_i \equiv \sum_{k=1}^{n^{(2)}} \Big|\omega^{(2)}_{ki}\Big| \, ,
\qquad i=1,\ldots,N_{\rm var} \, ,
\ee
with $\omega^{(2)}_{ki}$ the value of the weight connecting
the $k$-th neutron of the second layer with the $i$-th neuron of
the first (input) layer, and $n^{(2)}=5$ the number of
neurons in the second layer.
%
The input variables with a larger value of $\omega^{\rm tot}_i$ will those
be the ones that play a dominant role in enhancing the signal
discrimination using the MVA.

%
In Fig.~\ref{fig:nnweights} we show
the distribution of the total associated weight,
Eq.~(\ref{eq:totweight}) for each of the $N_{\rm var}$ input
variables of the resolved (left plot) and boosted (right plot) categories.
%
The notation for the various kinematic variables is the same
as in Sect.~\ref{sec:input}.
%
The important information
is contained in the relative strengths of the total associated weight
for each of the input variables.
%
The results of Fig.~\ref{fig:nnweights} illustrate which variables
are more discriminative between the signal and the background.
%
For instance, in the 
resolved category, the variables that lead to
a higher discrimination power
are the $p_T$ of the two reconstructed Higgs candidates, and then on a similar
footing other variables like the $p_T$ of the individual jets
and the mass of the di-Higgs system $m_{hh}$.
%
The only variable which seems to contain to information at all
is the $\Delta \phi_{hh}$ separation between the two
Higgs candidates.

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
\begin{center}
\includegraphics[width=0.49\textwidth]{plots/res_wgthist_noPU.pdf}
\includegraphics[width=0.49\textwidth]{plots/bst_wgthist_noPU.pdf}
\vspace{-0.5cm}
\caption{\small
Distribution of the total associated weight,
Eq.~(\ref{eq:totweight}) for each of the $N_{\rm var}$ input
variables of the resolved (left plot) and boosted (right plot) categories.
%
See Sect.~\ref{sec:input} for a description of each of the input
variables to the MVA.
}
\label{fig:nnweights}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

At this point we have all the ingredients that we need to determine the optimal
value of the MVA output cuts $y_{\rm cut}$
and quantify the final values of the signal
significance $S/\sqrt{B}$ and $S/B$ in the three categories.
%
These optimal values are determined from the maximisation of $S/\sqrt{B}$,
ensuring that the number of events $N_{\rm ev}$
which would be expected at the HL-LHC is large
enough.
%
In addition, we also require when fixing $y_{\rm cut}$
that the actual number of MC events used for the ANN training
is large enough to avoid the biases of a small training sample.

@@

The results of Fig.~\ref{fig:sb_mva}  show that
for all categories the  improvement in signal
significance arising from the MVA is substantial, growing from around 1
to around 5 in the three cases.
%
The optimal values of $y_{\rm cut}$ ensure that the
number of events at the HL-LHC is still well appreciable: we obtain
1300, 360 and 800 events in the resolved, intermediate, and
boosted categories respectively.
%
Another important benefit of the MVA applied to the boosted
category is that fact that the background is substantially
reduced so that $S/B$ can be as large as 10\% for the optimal
values of the ANN output cut.
%
Note that this is not true for the other two categories,
where $S/B$ is roughly constant.
%
A reasonably large value of $S/B$ is beneficial
from the experimental point of view, since it indicates
that the measurement is feasible even if the
systematic uncertainties are not very small.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
  \begin{center}
      \vspace{-1cm}
  \includegraphics[width=0.90\textwidth]{plots/bst_nnarch_noPU.pdf}
  \vspace{-2cm}
  \caption{\small Architecture of the Artificial
    Neural Network (ANN)
    used for the analysis of the
    boosted
    category, with $N_{\rm var}=21$ input variables and thus
    the same number of neurons
  in the first layer.
  %
  The color code in the neuron connections (the weights) is a heat map obtained
  at the end of the GA training,
  with red indicating larger values and black smaller values.
}
\label{fig:nnarch}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

The training of the ANN for the signal/background classification task
proceeds as follows.
%
Given a set of $N_{\mathrm{var}}$  kinematic variables $\{k\}_i$ associated with the event $i$, and a set of neural network parameters $\{\omega\}$, we interpret the neural network output $y_i$ (that is, the activation state of the
neuron in the last layer)
as the probability that the event $i$ originates from the signal process,
\be
y_i = P(y^\prime_i=1|\{k\}_i, \{\omega\} )\, ,
\ee
where $y_i^\prime$ represents the true classification of the event $i$, {\it i.e},
$y^\prime = 1$ for signal and $y^\prime = 0$ for background events.
%
With this interpretation, our general classification probability including background events is given by
\be
P(y_i^\prime|\{k\}_i, \{\omega\}) = y_i^{y^\prime_i}(1-y_i)^{1-y^\prime_i} \, ,
\ee
which implies that the  negative log-likelihood error function $E(\{\omega\})$
that needs to be minimized during the ANN training is 
the cross-entropy error, defined as
 \bea
 &&E(\{\omega\}) = -\log\left(\prod_i^{N_{\text{evt}}} P(y_i^\prime|\{k\}_i, \{\omega\})\right)\nonumber\\
 &&=
 \sum_i^{N_{\text{evt}}} y^\prime_i\log{y_i} + (1-y^\prime_i)\log{(1-y_i)} \, ,
 \label{cross-entropy}
 \eea
 where $N_{\text{evt}}$ is the number of
 Monte Carlo events that we have generated.
 %
 The ANN is trained both on the signal and background MC events,
 in each case using the information $y_i'$ of the true nature
 of the event.
 %
 To be stable against statistical fluctuations, it is important to use
 a large enough MC sample both of signal and of background events
 for the MVA training.

 Therefore, the training of the neural networks consist on the
 minimization of the cross-entropy error,
 Eq.~(\ref{cross-entropy}), which in this work is achieved using a
 Genetic Algorithm (GA).
 %
 The GA training is performed for a very large 
 number of generations, in this case $N_{\rm gen}=50k$, to avoid
 under-training.
 %
 We have verified that if a larger number of generations
 are used the results are unchanged.
 %

@@

 To identify the variables with higher discriminatory power,
 we need to train
 the MVA using a large number of kinematic distributions for
 signal and background events.
%
In this work we use different sets of
input variables for the MVA for the three categories.
%
In particular, in the case of large-$R$ jets (for the boosted
and intermediate categories) we are exploiting the jet
substructure information.

For the three categories, boosted, intermediate and resolved,
the following common variables are used as input to the MVA:
\begin{itemize}
\item The transverse momenta of the leading and subleading Higgs, $p_{T,h_1}$ and $p_{T,h_2}$.
\item The transverse momentum of the reconstructed Higgs pair, $p_{T,hh}$.
\item The invariant masses of the leading and sub-leading Higgs candidates, $m_{h,1}$ and $m_{h,2}$.
\item The invariant mass of the reconstructed Higgs pair, $m_{hh}$.
\item The separation in the $\phi$--$\eta$ plane
  between the two Higgs candidates, $\Delta R_{hh}$.
  \item The separation in $\eta$  between the two Higgs candidates, $\Delta \eta_{hh}$.
\item The separation in $\phi$  between the two Higgs candidates, $\Delta \phi_{hh}$.
\end{itemize}
In addition, in the boosted category we use
  the transverse momenta of the leading, $p_{T,h_{1,1}}$ and $p_{T,h_{1,2}}$ and
  sub-leading, $p_{T,h_{2,1}}$ and $p_{T,h_{2,1}}$, Higgs candidate subjets.
  %
  In the resolved category instead,
  the corresponding variables are
  the transverse momenta $p_{T,i}$ of the four leading 
  $b$-tagged small-$R$ jets in the event.
  %
  In the intermediate category, we use the
  transverse momenta of the subjets
  from the large-$R$ jet $p_{T,h_{1,1}}$ and $p_{T,h_{1,2}}$ and the
 transverse momenta $p_{T,i}$ of the two leading 
  $b$-tagged small-$R$ jets.
  %
 Therefore, we have 13 variables which are common to the three categories,
 and a total of $N_{\mathrm{var}}=13,17$ and 21 variables for the
resolvedm intermediate and boosted categories.

Given that the MVA is able to identify the most discriminatory variables
and to ignore those that have smaller effects, it is advantageous to
include as many input variables as possible.
%
This is one of the main advantages of ANNs in this context: they are
inherently redundant, so
adding additional information, even if carries very little weight,
should not degrade
the classification power of the MVA, other than making the GA
minimization less efficient.

@@

In Fig.~\ref{fig:nnresponse} we show the distribution of
the ANN output at the end of the training, for the exclusive
boosted, intermediate and resolved categories.
%
The distributions are normalized so that their integral
  adds up to one.
%
The  separation between signal and background is achieved by introducing
a cut, $y_{\rm cut}$, on the ANN output, so that MC events with $y_i\ge
y_{\rm cut}$ are classified as signal events, and those with
 $y_i <
y_{\rm cut}$ as background events.
%
Therefore,
the more separated that the two distributions are, the more efficient
the MVA discrimination will be.

@@

As can be seen from Fig.~\ref{fig:nev2}, in the boosted category
events, with a cut in the ANN output as hard as $y_{\rm cut}\simeq 0.9$,
we are still left
with around 100 events at the HL-LHC, 
which the background being reduced substantially to around
1000 events.
%
A similar cut would work fine for the intermediate category, though here
the backgrounds are much larger and thus the expected signal significance smaller.
%
For the resolved case, the signal significance would be only slightly
improved for any value of the cut.

A useful property of MVAs such as the one used in this work
is that they provide direct  physical insight about which of the
input variables contribute the most to the separation of
signal over background.
%
In the case of ANNs, this can be quantified by computing the sum
of the absolute value of al the weights connected to a given
input neuron $i$, that is
\be
\label{eq:totweight}
\omega^{\rm tot}_i \equiv \sum_{k=1}^{n^{(2)}} \Big|\omega^{(2)}_{ki}\Big| \, ,
\qquad i=1,\ldots,N_{\rm var} \, ,
\ee
with $\omega^{(2)}_{ki}$ the value of the weight connecting
the $k$-th neutron of the second layer with the $i$-th neuron of
the first (input) layer, and $n^{(2)}=5$ the number of
neurons in the second layer.
%
The input variables with a larger value of $\omega^{\rm tot}_i$ will be those
that play a dominant role in enhancing the signal
discrimination using the MVA.

%
In Fig.~\ref{fig:nnweights} we show
the distribution of the total associated weight,
Eq.~(\ref{eq:totweight}) for each of the $N_{\rm var}$ input
variables of the resolved (left plot) and boosted (right plot) categories.
%
The notation for the various kinematic variables is the same
as in Sect.~\ref{sec:input}.
%
The important information
is contained in the relative strengths of the total associated weight
for each of the input variables.
%
The results of Fig.~\ref{fig:nnweights} illustrate which variables
are more discriminative between the signal and the background.
%
For instance, in the 
resolved category, the variables that lead to
a higher discrimination power
are the $p_T$ of the two reconstructed Higgs candidates, and then on a similar
footing other variables like the $p_T$ of the individual jets
and the mass of the di-Higgs system $m_{hh}$.
%
The only variable which seems to contain no information at all
is the $\Delta \phi_{hh}$ separation between the two
Higgs candidates.

%%%%%%%%%%%%%%%%%%%%%%%%
\begin{figure}[t]
  \begin{center}
    \includegraphics[width=0.49\textwidth]{plots/res_wgthist_noPU.pdf}
\includegraphics[width=0.49\textwidth]{plots/int_wgthist_noPU.pdf}
\includegraphics[width=0.75\textwidth]{plots/bst_wgthist_noPU.pdf}
\vspace{-0.5cm}
\caption{\small
Distribution of the total associated weight,
Eq.~(\ref{eq:totweight}) for each of the $N_{\rm var}$ input
variables of the resolved (upper left plot),  intermediate (upper right plot)
and boosted (lower plot)
categories.
%
See Sect.~\ref{sec:input} for a description of each of the input
variables to the MVA.
}
\label{fig:nnweights}
\end{center}
\end{figure}
%%%%%%%%%%%%%%%%%%%%%%%

@@

At this point we have all the
information needed to determine the optimal
value of the MVA output cuts $y_{\rm cut}$
and quantify the final values of the signal
significance $S/\sqrt{B}$ and $S/B$ in the three categories.
%
These optimal values are determined from the maximisation of $S/\sqrt{B}$,
ensuring that the number of events $N_{\rm ev}$
which would be expected at the HL-LHC is large
enough.
%
In addition, we also require when fixing $y_{\rm cut}$
that the actual number of MC events used for the ANN training
is large enough to avoid the biases of a small training sample.

@@

The results of Fig.~\ref{fig:sb_mva}  show that
for all categories the  improvement in signal
significance as compared to the pre-MVA results
is substantial, growing from around 1
to around 5 in the three cases.
%
The optimal values of $y_{\rm cut}$ ensure that the
number of events at the HL-LHC is still well appreciable: we obtain
1300, 360 and 800 events in the resolved, intermediate, and
boosted categories respectively.
%
Another important benefit of the MVA applied to the boosted
category is that fact that the background is substantially
reduced so that $S/B$ can be as large as 10\% for the optimal
values of the ANN output cut.
%
Note that this is not true for the other two categories,
where $S/B$ is roughly constant.
%
A reasonably large value of $S/B$ is beneficial
from the experimental point of view, since it indicates
that the measurement is feasible even if the
systematic uncertainties are not very small.
